---
title: "Exploratory Analysis Script"
author: "Taylor Glass and Arlyn Santiago"
date: "02/21/24"
output: html_document
---

# Setup

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
```

Load the data.

```{r}
#Path to data
data_location <- here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```


# Data exploration through tables
We will begin with creating a summary table of our cleaned variables using the skim()function. This provides information about each of the numeric variables including the mean, standard deviation, quartiles, missing observations, and a histogram. 

```{r}
# skim the data to create summary file
summary_df = skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```
We also created simple tables for the categorical variables: `region`, `larcster_usage`, `sarc_usage`, and `none_usage`.
```{r}

```


# Data exploration through figures

We will make histogram plots for the continuous outcomes in our dataset.

It is important to consider the type of birth control methods associated with percent of unintended pregnancies as well. We will create a histogram for the `pct_upreg_larcster`, `pct_upreg_sarc`, `pct_upreg_trad`, and `pct_upreg_nouse` variables.  

```{r}
# explore the percentage of unintended pregnancies related to long acting reversible methods in each country
p2 <- mydata %>% ggplot(aes(x=pct_upreg_larcster)) + geom_histogram() 
plot(p2)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_larcster_distribution.png")
ggsave(filename = figure_file, plot=p2) 

# explore the percentage of unintended pregnancies related to short acting reversible methods in each country
p3 <- mydata %>% ggplot(aes(x=pct_upreg_sarc)) + geom_histogram() 
plot(p3)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_sarc_distribution.png")
ggsave(filename = figure_file, plot=p3) 

# explore the percentage of unintended pregnancies related to traditional methods in each country
p4 <- mydata %>% ggplot(aes(x=pct_upreg_trad)) + geom_histogram() 
plot(p4)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_trad_distribution.png")
ggsave(filename = figure_file, plot=p4) 

# explore the percentage of unintended pregnancies related to no birth control usage in each country
p5 <- mydata %>% ggplot(aes(x=pct_upreg_nouse)) + geom_histogram() 
plot(p5)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_nouse_distribution.png")
ggsave(filename = figure_file, plot=p5) 
```

We will now explore the new total usage variables that were created using the number of women using each method across the four age categories. This set of plots revealed large outliers in the 'total_larcster`, 'total_sarc', 'total_trad', and `total_none` variables, which prevented the graphs from being very useful. The histograms look like most observations for each of these four variables are 0, but that is not possible because the summary table showed that the minimum value for all four of these variables is greater than 0. This distortion is caused by the two large outliers in the data set.    

```{r}
# explore the total number of people using long acting reversible methods in each country
p6 <- mydata %>% ggplot(aes(x=total_larcster)) + geom_histogram() 
plot(p6)
figure_file = here("results", "figures", "exploratory-figures", "total_larcster_distribution.png")
ggsave(filename = figure_file, plot=p6) 

# explore the total number of people using short acting reversible methods in each country
p7 <- mydata %>% ggplot(aes(x=total_sarc)) + geom_histogram() 
plot(p7)
figure_file = here("results", "figures", "exploratory-figures", "total_sarc_distribution.png")
ggsave(filename = figure_file, plot=p7) 

# explore the total number of people using traditional methods in each country
p8 <- mydata %>% ggplot(aes(x=total_trad)) + geom_histogram() 
plot(p8)
figure_file = here("results", "figures", "exploratory-figures", "total_trad_distribution.png")
ggsave(filename = figure_file, plot=p8) 

# explore the total number of people not using birth control in each country
p9 <- mydata %>% ggplot(aes(x=total_none)) + geom_histogram() 
plot(p9)
figure_file = here("results", "figures", "exploratory-figures", "total_none_distribution.png")
ggsave(filename = figure_file, plot=p9) 
```

We need to explore where the outliers in these variables are coming from. China and India are the two countries that account for the outliers in all 4 variables, which is likely being skewed by population size. The first method of standardizing we tried is creating proportions using the total count for each birth control method type. After creating these standardized variables, the same issue with outliers remained because China and India still account for the largest proportion of observations. Next, we considered using a scaling factor. We attempted this approach with the `total_larcster` variable and found the same issues as before. The third attempt at standarization utilized a min-max scaling with z-score normalization, but the outlier issue remained. 
```{r}
# find the countries that represent the outliers
outliers <- mydata %>% filter(total_larcster > 25000, 
                              total_sarc > 25000,
                              total_trad > 5000,
                              total_none > 10000)

# standardization attempt 1
mydata <- mydata %>% mutate(larcster_st = total_larcster / sum(total_larcster),
                            sarc_st = total_sarc / sum(total_sarc),
                            trad_st = total_trad / sum(total_trad),
                            none_st = total_none / sum(total_none))

# standardization attempt 2
scaling_factor_larcster <- max(mydata$total_larcster)
mydata <- mydata %>% mutate(standardized_larcster = total_larcster / scaling_factor_larcster)

# standardization attempt 3
min_count <- min(mydata$total_larcster)
max_count <- max(mydata$total_larcster)
scaled_counts <- (mydata$total_larcster - min_count) / (max_count - min_count)
print(scaled_counts)
mean_count <- mean(mydata$total_larcster)
sd_count <- sd(mydata$total_larcster)
z_score_counts <- (mydata$total_larcster - mean_count) / sd_count
```

After reviewing the raw data set again, we found the `wra` variable that accounts for the number of women of reproductive age (15-49) in each country. Once this variable was added to the original processed data set, we could use it to standardize the total counts for each birth control method by population size of women of reprodutive age. 
```{r}
# population size standarization for total counts per birth control method
mydata <- mydata %>% mutate(larcster_standard = total_larcster / wra,
                            sarc_standard = total_sarc / wra,
                            trad_standard = total_trad / wra, 
                            none_standard = total_none / wra)
```

We will start exploring how many women use each type of birth control in each country compared to the percent of unintended pregnancies related to that birth control. Plot 10 shows `pct_upreg_larcster` vs `larcster_standard` to analyze women using long acting reversible methods and sterilization versus the percentage of unintended pregnancies using those methods. Plot 11 uses `pct_upreg_sarc` and `sarc_standard` to examine the same relationship as plot 10 but with short-acting reversible methods. Plot 12 shows `pct_upreg_trad` versus `trad_standard`, and plot 13 shows `pct_upreg_nouse` versus `none_standard`. 
```{r}
# explore percentage of unintended pregnancies compared to standardized count of women using long-acting reversible and sterilization methods
p10 <- mydata %>% ggplot(aes(x=pct_upreg_larcster, y=larcster_standard)) + geom_point() + geom_smooth(method='lm')
plot(p10)
figure_file = here("results", "figures", "exploratory-figures", "pctuppreg_totallarcster.png")
ggsave(filename = figure_file, plot=p10) 

# explore percentage of unintended pregnancies compared to standardized count of women using short acting reversible methods
p11 <- mydata %>% ggplot(aes(x=pct_upreg_sarc, y=sarc_standard)) + geom_point() + geom_smooth(method='lm')
plot(p11)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_totalsarc.png")
ggsave(filename = figure_file, plot=p11) 

# explore percentage of unintended pregnancies compared to standardized count of women using traditional methods
p12 <- mydata %>% ggplot(aes(x=pct_upreg_trad, y=trad_standard)) + geom_point() + geom_smooth(method='lm')
plot(p12)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_totaltrad.png")
ggsave(filename = figure_file, plot=p12) 

# explore percentage of unintended pregnancies compared to standardized count of women not using no method
p13 <- mydata %>% ggplot(aes(x=pct_upreg_nouse, y=none_standard)) + geom_point() + geom_smooth(method='lm')
plot(p13)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_totalnone.png")
ggsave(filename = figure_file, plot=p13) 
```

We are also interested in how these relationships between percent of unintended pregnancies using each birth control method and the number of people using each method varies across region. We adjusted the original scatter plots created above by coloring based on region. 

```{r}
# explore percent unintended pregnancies using long acting reversible and sterilziation methods by region
p14 <- mydata %>% ggplot(aes(x=pct_upreg_larcster, y=larcster_standard, color = region)) + geom_point() + geom_smooth(method='lm')
plot(p14)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_larcster_stratified.png")
ggsave(filename = figure_file, plot=p14) 

# explore percent unintended pregnancies using short acting reversible methods by region
p15 <- mydata %>% ggplot(aes(x=pct_upreg_sarc, y=sarc_standard, color = region)) + geom_point() + geom_smooth(method='lm')
plot(p15)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_sarc_stratified.png")
ggsave(filename = figure_file, plot=p15) 

# explore percent unintended pregnancies using traditional methods by region
p16 <- mydata %>% ggplot(aes(x=pct_upreg_trad, y=trad_standard, color = region)) + geom_point() + geom_smooth(method='lm')
plot(p16)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_trad_stratified.png")
ggsave(filename = figure_file, plot=p16) 

# explore percent unintended pregnancies using no method by region
p17 <- mydata %>% ggplot(aes(x=pct_upreg_nouse, y=none_standard, color = region)) + geom_point() + geom_smooth(method='lm')
plot(p17)
figure_file = here("results", "figures", "exploratory-figures", "pctupreg_none_stratified.png")
ggsave(filename = figure_file, plot=p17) 
```

Our next group of plots will be violin plots to explore the usage level of each birth control method, a categorical variable, associated with percentage of unintended pregnancies use for each birth control method, a numerical variable.
```{r}
# explore long acting reversible and sterilization method usage compared to percent of unintended pregnancies
p18 <- mydata %>% ggplot(aes(x = larcster_usage, y = pct_upreg_larcster)) +
  geom_violin() +
  labs(x = "Long Acting Reversible and Sterilization Usage", y = "Percentage Unintended Pregnancies") 
plot(p18)
figure_file = here("results", "figures", "exploratory-figures", "larcster_pctupreg_violin.png")
ggsave(filename = figure_file, plot=p18) 

# explore short acting reversible method usage compared to percent of unintended pregnancies
p19 <- mydata %>% ggplot(aes(x = sarc_usage, y = pct_upreg_sarc)) +
  geom_violin() +
  labs(x = "Short Acting Reversible Usage", y = "Percentage Unintended Pregnancies") 
plot(p19)
figure_file = here("results", "figures", "exploratory-figures", "sarc_pctupreg_violin.png")
ggsave(filename = figure_file, plot=p19) 

# explore traditional method usage compared to percent of unintended pregnancies
p20 <- mydata %>% ggplot(aes(x = trad_usage, y = pct_upreg_trad)) +
  geom_violin() +
  labs(x = "Traditional Method Usage", y = "Percentage Unintended Pregnancies") 
plot(p20)
figure_file = here("results", "figures", "exploratory-figures", "trad_pctupreg_violin.png")
ggsave(filename = figure_file, plot=p20) 

# explore no method usage compared to percent of unintended pregnancies
p21 <- mydata %>% ggplot(aes(x = none_usage, y = pct_upreg_nouse)) +
  geom_violin() +
  labs(x = "No Birth Control Usage", y = "Percentage Unintended Pregnancies") 
plot(p21)
figure_file = here("results", "figures", "exploratory-figures", "none_pctupreg_violin.png")
ggsave(filename = figure_file, plot=p21) 
```
