---
title: "Processing Script"
author: "Taylor Glass and Arlyn Santiago"
date: "2024-02-16"
output: html_document
---


# Processing script

# Setup

Load the necessary packages for data loading and processing. 

```{r}
#| message: false
library(readr) # for loading csv files
library(readxl) # for loading excel files (codebook)
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(naniar) #to explore data missingness
library(skimr) #for nice visualization of data 
library(here) #to set paths
```


# Data loading

```{r}
#| message: false
# path to data using here() function
data_location <- here("data","raw-data","AIU All Women Dataset.csv")
rawdata <- read_csv(data_location)
```
There are 132 observations, one row for each low- to middle- income country, of the 399 variable included in the Adding It Up All Women Dataset. Some examples of numeric variables included are percentages of pregnancy outcomes, rates of safe and unsafe abortions, percentage of various contraceptive usages, percentages of care received for various pregnancy complications, and costs for many variables including abortions and STIs. There are three important character variables that can be used to group the numeric variables: country name, region, and sub region. 

# Check data

We will load and examine the codebook to learn more about the dataset. 

```{r}
#| message: false
# write file path to the codebook
codebook_location <- here("data", "AIU All Women Dataset Codebook.xlsx")
  
# specify the 3 sheets of the codebook we need 
sheets <- c("Demographics", "Contraceptive need & use", "Pregnancy-related & newborn")

# load codebook
codebook <- lapply(sheets, function(sheet) {
  readxl::read_excel(codebook_location, sheet = sheet)
})
print(codebook)
```
Note: should we include other covariates like marriage percentages, number of live births, number of induced abortions

# Glimpse at the data

We will take a glimpse at our data to better understand the variables we are working with. There are four character variables: `country`, `data_source_year`, `region`, and `subregion`. The remainder of the variables are numeric and describe percentage of birth outcomes, maternal deaths due to birth outcomes, contraceptive methods divided by age groups, wealth levels in relation to contraceptive methods, various abortion procedure rates, current cost of pregnancy outcomes and STI treatment.

```{r}
#| message: false
dplyr::glimpse(rawdata)
summary(rawdata)
head(rawdata)
skimr::skim(rawdata)
```

# Identifying variables to be cleaned 
We want to explore the relationship between unintended pregnancies and contraception methods. We will use three predictor variables: total number of women using the contraceptive method, region, number of safe abortions,  . The variable for the total percentage of unintended pregnancies, `pct_upreg` will be the main outcome variable with additional options for outcome variables of total percentage of unintended pregnancies by each contraceptive method: long acting reversible methods (`pct_upreg_larcster`), short term reversible methods (`pct_upreg_sarc`), traditional methods (`pct_upreg_trad`), and no method (`pct_upreg_nouse`). We will use multiple predictor variables including: total number of women using the contraceptive method, region, subregion, percentage of married women, percentage of women never married, total maternal deaths per 100,000 live births, deaths from unintended pregnancy per 100,000 live births, and deaths related to abortion per 100000 abortions. Covariates will include the number of women using each of these methods across four different age groups: 15-19, 20-24, 25-34, and 35-49. These variables will be summed to create total number of women using each type of contraception: 
-Number of women 15-19 using long-acting reversible methods `larc_1519`
-Number of women 15-19 using short-acting methods `sarc_1519`
-Number of women 15-19 using traditional methods `tradmethods_1519`
-Number of women 15-19 using no method `nomethod_1519`
-Number of women 20-24 using long-acting reversible methods `larc_2024`
-Number of women 20-24 using short-acting methods	`sarc_2024`
-Number of women 20-24 using traditional methods	`tradmethods_2024`
-Number of women 20-24 using no method	`nomethod_2024`
-Number of women 25-34 using long-acting reversible methods	`larc_2534`
-Number of women 25-34 using short-acting methods	`sarc_2534`
-Number of women 25-34 using traditional methods	`tradmethods_2534`
-Number of women 25-34 using no method	`nomethod_2534`
-Number of women 35-49 using long-acting reversible method	`larc_3549`
-Number of women 35-49 using short-acting method	`sarc_3549`
-Number of women 35-49 using all traditional methods	`tradmethods_3549`
-Number of women 35-49 using no method	`nomethod_3549`



```{r}
#| message: false
# select the outcome variable, predictor variables, and covariates  
rawdata2 <- rawdata %>% 
                select(country, wra, region, subregion, pct_currentlymarried, pct_nevermarried, rate_matdeaths, rate_matdeaths_upreg, rate_matdeaths_abs, pct_upreg, upreg_subreg, pct_upreg_larcster, pct_upreg_sarc, pct_upreg_trad, pct_upreg_nouse, larc_1519, ster_1519, sarc_1519, tradmethods_1519, nomethod_1519, larc_2024, ster_2024, sarc_2024, tradmethods_2024, nomethod_2024, larc_2534, ster_2534, sarc_2534, tradmethods_2534,  nomethod_2534, larc_3549, ster_3549, sarc_3549, tradmethods_3549, nomethod_3549)

# take a glimpse at the selected variables 
skim(rawdata2)
head(rawdata2)
```

# Cleaning

By inspecting the data as done above, we find some problems that need addressing:

First, the `region` variable is a character variable, but it will be more useful later as a factor. We can easily mutate it from a character variable to factor variable. All the remaining predictor and covariate variables are numeric as they should be. 
```{r}
#| message: false
# convert character variable to factor variable
rawdata2 <- rawdata2 %>% 
              mutate(region = as.factor(region))

# check that the conversion worked 
class(rawdata2$region)
```

Second, there are 24 entries missing for the percentage of all unintended pregnancies by country `pct_upreg`, which accounts for 18.18% of the observations. After isolating the observations with this missing variable, we see that the majority of countries in this category are located in Asia. This does not make sense because the majority of countries in the raw dataset are located in Africa, which suggests that the data are not missing at random. 
```{r}
#| message: false
# find missing percentage for each variable
miss_var_summary(rawdata2)

# isolate the percentage of all unintended pregnancies by country to discover trends in missingness 
variable_with_missing <- "pct_upreg"
missing_obs <- rawdata2[is.na(rawdata2[[variable_with_missing]]), ]
other_variables <- colnames(rawdata2)[!colnames(rawdata2) == variable_with_missing]
pct_upregmissing <- missing_obs[, other_variables]
print(pct_upregmissing)

# compare dataset with the missing variable to the full dataset
table(pct_upregmissing$region)
table(rawdata2$region)
```

We also need to evaluate the 108 missing entries of the percentage of unintended pregnancies by subregion `upreg_subreg`. Since 81.81% of this variable is missing, it does not make sense to include it in the analysis because we do not have enough information to impute the missing data. We will remove this variable for the sake of this analysis.
```{r}
#| message: false
# isolate the percentage of all unintended pregnancies by subregion to discover trends in missingness
variable_with_missing <- "upreg_subreg"
missing_obs2 <- rawdata2[is.na(rawdata2[[variable_with_missing]]), ]
other_variables2 <- colnames(rawdata2)[!colnames(rawdata2) == variable_with_missing]
pct_subregmissing <- missing_obs2[, other_variables2]
print(pct_subregmissing)

# remove unintended pregnancies by subregion variable due to missingness
rawdata2 <- rawdata2 %>% 
              select(-upreg_subreg)
```

Fortunately, the rest of the variables required for this analysis are complete including the `wra` variable with population size that we will need to standardize population counts later in the exploratory data analysis process. Each of the four variables describing the percentage of unintended pregnancies based on birth control type and the two variables described percentage of married and non-married women also need to be converted to traditional percentage format. 
```{r}
#| message: false
# convert percentage variable to typical percentage format
rawdata2 <- rawdata2 %>% 
              mutate(pct_upreg_larcster = pct_upreg_larcster * 100, 
                     pct_upreg_sarc = pct_upreg_sarc * 100, 
                     pct_upreg_trad = pct_upreg_trad * 100, 
                     pct_upreg_nouse = pct_upreg_nouse * 100,
                     pct_currentlymarried = pct_currentlymarried * 100,
                     pct_nevermarried = pct_nevermarried * 100)
  

# ensure mutations worked correctly
summary(rawdata2$pct_upreg_larcster)
summary(rawdata2$pct_upreg_sarc)
summary(rawdata2$pct_upreg_trad)
summary(rawdata2$pct_upreg_nouse)
summary(rawdata2$pct_currentlymarried)
summary(rawdata2$pct_nevermarried)
```

We should address the missing data in the general `pct_upreg` here to avoid introducing bias because 18.18% of this variable is missing, and it is essential for answering the research question. After reviewing the codebook, the researchers provided an external source for this variable, which we have included in the references. The numbers in this source do not match what is included in the data set. After exploring the methodology for the Adding It Up study, we found their original data that matches what is recorded in the dataset. When we tried to create our own percentage of unintended pregnancy variable, we found that each birth control method adds up to account for 100% of unintended pregnancies. Since there are 24 observations missing in what should be the total unintended pregnancy variable, it is likely that the variables with the percentages of unintended pregnancies by birth control method were imputed or estimated. Missingness in the collection of data could also explain why the long-acting reversible contraceptives and sterilization method were combined into one category. Based on this discovery, we modified the question that we want to answer with the data to use variables that are complete and reliable: `pct_upreg_larcster`, `pct_upreg_sarc`, `pct_upreg_trad`, and `pct_upreg_nouse`. These variables will allow us to answer the question of is there a significant difference in the percentage of unplanned pregnancies between three contraceptive methods among all countries compared to no usage. 

```{r}
#| message: false
rawdata2 <- rawdata2 %>% 
            mutate(pct_preg = (pct_upreg_larcster + pct_upreg_sarc + pct_upreg_trad + pct_upreg_nouse))
summary(rawdata2$pct_preg)
```

To make the covariate variables more useful for answering our research question, it makes sense to create a new factor variable denoting low, medium, and high usage of each birth control method by country. First, we need to determine the boundaries for each level of usage by exploring how many women use each method across age groups. We need to create a variable showing the total usage of each birth control method to remove the effect of age. It is worth noting again that the long acting reversible and sterilization methods are combined in the`pct_upreg_larcster` variable, so we need to consider all the women using both of these methods in the total. 
```{r}
#| message: false
# create new variables showing total usage of each method including all age groups
rawdata2 <- rawdata2 %>% 
              mutate(total_larcster = larc_1519 + larc_2024 + larc_2534 + larc_3549 + ster_1519 + ster_2024 + ster_2534 +                          ster_3549,
                     total_sarc = sarc_1519 + sarc_2024 + sarc_2534 + sarc_3549,
                     total_trad = tradmethods_1519 + tradmethods_2024 + tradmethods_2534 + tradmethods_3549,
                     total_none = nomethod_1519 + nomethod_2024 + nomethod_2534 + nomethod_3549)

# explore summary of each total to set boundaries for each level of usage
rawdata2 %>% skim(total_larcster, total_sarc, total_trad, total_none)
```

After finding the distribution of number of women using each type of birth control, we can use that information to create a new factor variable denoting low, medium, and high usage for each method. The "low" category will include observations below the 25th percentile. The "medium" category will include observations between the 25th and 75th percentiles. The "high" category will include observations above the 75th percentile. 
```{r}
#| message: false
# create usage variable for long acting reversible methods 
rawdata2 <- rawdata2 %>% 
              mutate(larcster_usage = as.factor(case_when(
                                  total_larcster < 23.94865 ~ "Low",
                                  total_larcster >= 23.94865 & total_larcster < 1554.7609 ~ "Medium",
                                  total_larcster >= 1554.7609 ~ "High")))

# check if new variable was created correctly 
class(rawdata2$larcster_usage)

# create usage variable for short acting reversible methods 
rawdata2 <- rawdata2 %>% 
              mutate(sarc_usage = as.factor(case_when(
                                  total_sarc < 114.33113 ~ "Low",
                                  total_sarc >= 114.33113 & total_sarc < 1554.7609 ~ "Medium",
                                  total_sarc >= 1554.7609	 ~ "High")))

# check if new variable was created correctly 
class(rawdata2$sarc_usage)

# create usage variable for traditional methods 
rawdata2 <- rawdata2 %>% 
              mutate(trad_usage = as.factor(case_when(
                                  total_trad < 15.05000	 ~ "Low",
                                  total_trad >= 15.05000 & total_trad < 332.4750 ~ "Medium",
                                  total_trad >= 332.4750	 ~ "High")))

# check if new variable was created correctly 
class(rawdata2$trad_usage)

# create usage variable for no methods
rawdata2 <- rawdata2 %>% 
              mutate(none_usage = as.factor(case_when(
                                  total_none < 80.07500	 ~ "Low",
                                  total_none >= 80.07500 & total_none < 875.2750 ~ "Medium",
                                  total_none >= 875.2750	 ~ "High")))

# check if new variable was created correctly 
class(rawdata2$none_usage)
```
The exploratory data analysis in the eda.qmd file in the code > eda-code folder revealed outliers in the data that need to be addressed. 

The two outliers make the histograms showing the distribution of total usage of birth control method by country very difficult to interpret. We need to explore where the outliers in these variables are coming from. 
```{r}
# find the countries that represent the outliers
outliers <- rawdata2 %>% filter(total_larcster > 25000, 
                              total_sarc > 25000,
                              total_trad > 5000,
                              total_none > 10000)

# print the outliers
outliers
```

China and India are the two countries that account for the outliers in all 4 variables, which is likely being skewed by population size. 

The first method of standardizing we tried is creating proportions using the total count for each birth control method type. After creating these standardized variables, the same issue with outliers remained because China and India still account for the largest proportion of observations. Next, we considered using a scaling factor. We attempted this approach with the `total_larcster` variable and found the same issues as before. The third attempt at standarization utilized a min-max scaling with z-score normalization, but the outlier issue remained. 
```{r}
# standardization attempt 1
rawdata2 <- rawdata2 %>% mutate(larcster_st = total_larcster / sum(total_larcster),
                            sarc_st = total_sarc / sum(total_sarc),
                            trad_st = total_trad / sum(total_trad),
                            none_st = total_none / sum(total_none))

# standardization attempt 2
scaling_factor_larcster <- max(rawdata2$total_larcster)
rawdata2 <- rawdata2 %>% mutate(standardized_larcster = total_larcster / scaling_factor_larcster)

# standardization attempt 3
min_count <- min(rawdata2$total_larcster)
max_count <- max(rawdata2$total_larcster)
scaled_counts <- (rawdata2$total_larcster - min_count) / (max_count - min_count)
print(scaled_counts)
mean_count <- mean(rawdata2$total_larcster)
sd_count <- sd(rawdata2$total_larcster)
z_score_counts <- (rawdata2$total_larcster - mean_count) / sd_count
```

After reviewing the raw data set again, we found the `wra` variable that accounts for the number of women of reproductive age (15-49) in each country. Once this variable was added to the original processed data set, we could use it to standardize the total counts for each birth control method by population size of women of reproductive age. 
```{r}
# population size standardization for total counts per birth control method
rawdata2 <- rawdata2 %>% mutate(larcster_standard = total_larcster / wra,
                            sarc_standard = total_sarc / wra,
                            trad_standard = total_trad / wra, 
                            none_standard = total_none / wra)
```
 

Lastly, we will assign the clean data to a final variable, which makes it easier to add further cleaning steps above if necessary.

```{r}
#| message: false
processeddata <- rawdata2
```


# Save data 

Finally, we will save the clean data as RDS file to preserve coding like factors, characters, numeric, etc. 

```{r}
#| message: false
save_data_location <- here("data","processed-data","processeddata.rds")
saveRDS(processeddata, file = save_data_location)
```


