---
title: "Statistical Analysis Script"
author: "Taylor Glass and Arlyn Santiago"
date: "03/12/24"
output: html_document
---
## Load necessary packages
```{r}
#| message: false
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidymodels) # for model building 
```

## Load the cleaned data
```{r}
#path to data
data_location <- here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

## Data fitting and statistical analysis 
The first trend we will investigate is if there is a stronger relationship between total usage of one method of birth control and percentage of unintended pregnancies associated with that method across the three birth control methods and no usage options. These models will allow us to predict unintended pregnancies based on population data about number of women using each type of contraception method per country. 

### First Model Fit 
The first model attempts to predict the percentage of unintended pregnancies among women using long acting reversible  and sterilization methods of contraception, 'pct_upreg_larcster', based on the total number of women using this method of contraception. We will use the version of the variable that has been standardized by population size, larcster_standard.
```{r}
# fit linear model using pct_upreg_larcster as outcome, larcster_standard as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit1 <- lm_mod %>% 
            fit(pct_upreg_larcster ~ larcster_standard, mydata) #estimate/train the model 

#generate clean output with estimates and p-values
tidy(lmfit1) 

# save results from fit into a data frame 
lmtable1 <- tidy(lmfit1)

# save fit results table  
table_file1 = here("results", "resulttable1.rds")
saveRDS(lmtable1, file = table_file1)
```

### Second Model Fit 
The second model attempts to predict the percentage of unintended pregnancies among women using short acting reversible methods of contraception 'pct_upreg_sarc' based on the total number of women using this method of contraception. We will use the version of the variable that has been standardized by population size, 'sarc_standard'. 
```{r}
# fit linear model using pct_upreg_sarc as outcome, sarc_standard as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit2 <- lm_mod %>% 
            fit(pct_upreg_sarc ~ sarc_standard, mydata) #estimate/train the model 

#generate clean output with estimates and p-values
tidy(lmfit2) 

# save results from fit into a data frame 
lmtable2 <- tidy(lmfit2)

# save fit results table  
table_file2 = here("results", "resulttable2.rds")
saveRDS(lmtable2, file = table_file2)
```
### Third Model Fit 
The second model attempts to predict the percentage of unintended pregnancies among women using traditional of contraception 'pct_upreg_trad' based on the total number of women using this method of contraception. We will use the version of the variable that has been standardized by population size, 'trad_standard'. 
```{r}
# fit linear model using pct_upreg_trad as outcome, trad_standard as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit3 <- lm_mod %>% 
            fit(pct_upreg_trad ~ trad_standard, mydata) #estimate/train the model 

# generate clean output with estimates and p-values
tidy(lmfit3)

# save results from fit into a data frame 
lmtable3 <- tidy(lmfit3)

# save fit results table  
table_file3 = here("results", "resulttable3.rds")
saveRDS(lmtable3, file = table_file3)
```
### Fourth Model Fit 
The fourth model attempts to predict the percentage of unintended pregnancies among women using no method of contraception 'pct_upreg_nouse' based on the total number of women not using a method of contraception. We will use the version of the variable that has been standardized by population size, 'none_standard'. 
```{r}
# fit linear model using pct_upreg_nouse as outcome, none_standard as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit4 <- lm_mod %>% 
            fit(pct_upreg_nouse ~ none_standard, mydata) #estimate/train the model 

# generate clean output with estimates and p-values
tidy(lmfit4) 

# save results from fit into a data frame 
lmtable4 <- tidy(lmfit4)

# save fit results table  
table_file4 = here("results", "resulttable4.rds")
saveRDS(lmtable4, file = table_file4)
```
Based on the coefficient estimates for each parameter, the association between total number of women not using a contraception method is most strongly correlated with percentage of unintended pregnancies among those not using contraceptive methods, followed by those who use traditional methods and short-acting reversible methods. The long acting reversible and sterilization methods shows the weakest association based on these four models. 


We want to explore if a model with all the variables of total number of women using each type of contraceptive method would be more accurate in estimating percentage of unintended pregnancies. 

### Fifth Model Fit 
```{r}
# fit linear model using pct_upreg_nouse as outcome, none_standard as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit5 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_larcster ~ larcster_standard + sarc_standard + trad_standard + none_standard, mydata)      

#generate clean output with estimates and p-values
tidy(lmfit5) 

# save results from fit into a data frame 
lmtable5 <- tidy(lmfit5)

# save fit results table  
table_file5 = here("results", "resulttable5.rds")
saveRDS(lmtable5, file = table_file5)
```
These results show that only 'larcster_standard' showed a positive relationship with percentage of unintended pregnancies associated with long acting reversible and standardization techniques. This is to be expected. 

We also want to determine if there is a relationship between the region and percentage of unintended pregnancies associated with each method of contraception. 

### Sixth Model Fit 
```{r}
# fit linear model using pct_upreg_nouse as outcome, none_standard as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit6 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_larcster ~ region, mydata)      

#generate clean output with estimates and p-values
tidy(lmfit6) 

# save results from fit into a data frame 
lmtable6 <- tidy(lmfit6)

# save fit results table  
table_file6 = here("results", "resulttable6.rds")
saveRDS(lmtable6, file = table_file6)
```
### Seventh Model Fit 
```{r}
# fit linear model using pct_upreg_sarc as outcome, pct_currentlymarried as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit7 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_sarc ~ pct_currentlymarried, mydata)      

#generate clean output with estimates and p-values
tidy(lmfit7) 

# save results from fit into a data frame 
lmtable7 <- tidy(lmfit7)

# save fit results table  
table_file7 = here("results", "resulttable7.rds")
saveRDS(lmtable7, file = table_file7)
```

# Eighth Model Fit 
```{r}
# fit linear model using pct_upreg_sarc as outcome, pct_nevermarried as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit8 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_sarc ~ pct_nevermarried, mydata)      

#generate clean output with estimates and p-values
tidy(lmfit8) 

# save results from fit into a data frame 
lmtable8 <- tidy(lmfit8)

# save fit results table  
table_file8 = here("results", "resulttable7.rds")
saveRDS(lmtable8, file = table_file8)
```

# Ninth Model Fit 
```{r}
# fit linear model using pct_upreg_sarc as outcome, rate_matdeaths as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit9 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_sarc ~ rate_matdeaths, mydata)      

#generate clean output with estimates and p-values
tidy(lmfit9) 

# save results from fit into a data frame 
lmtable9 <- tidy(lmfit9)

# save fit results table  
table_file9 = here("results", "resulttable9.rds")
saveRDS(lmtable9, file = table_file9)
```

# Tenth Model Fit 
```{r}
# fit linear model using pct_upreg_sarc as outcome, rate_matdeaths_upreg as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit10 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_sarc ~ rate_matdeaths_upreg, mydata)      

#generate clean output with estimates and p-values
tidy(lmfit10) 

# save results from fit into a data frame 
lmtable10 <- tidy(lmfit10)

# save fit results table  
table_file10 = here("results", "resulttable10.rds")
saveRDS(lmtable10, file = table_file10)
```

# Eleventh Model Fit 
```{r}
# fit linear model using pct_upreg_sarc as outcome, rate_matdeaths_abs as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit11 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_sarc ~ rate_matdeaths_abs, mydata)      

#generate clean output with estimates and p-values
tidy(lmfit11) 

# save results from fit into a data frame 
lmtable11 <- tidy(lmfit11)

# save fit results table  
table_file11 = here("results", "resulttable9.rds")
saveRDS(lmtable11, file = table_file11)
```
